---
title: "Anatomy of a Hit"
subtitle: "Interactive exploration of song length, danceability, tempo, and acousticness across 5,800+ Billboard hits."
format:
  html:
    toc: true
    page-layout: article
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(plotly)
library(reactable)
library(htmltools)

songs <- read_csv("data/billboard_audio_features.csv", show_col_types = FALSE)

genre_colors <- c(
  "Rock" = "#ef4444",
  "Pop" = "#ec4899",
  "Hip-Hop/Rap" = "#2563eb",
  "R&B/Soul" = "#8b5cf6",
  "Country" = "#f59e0b",
  "Electronic/Dance" = "#06b6d4"
)
```

## The Shrinking Song: Duration Across Decades

Why are modern songs getting shorter? In the 1960s, 7-inch vinyl records limited songs to under 3 minutes. The CD revolution in the 1980s and 1990s unlocked epic 5-minute anthems. But since 2015, streaming royalty models (which pay per play once a listener passes 30 seconds) and short-form video on TikTok have pushed hit songs back down to historic lows.

```{r}
#| label: song-duration-chart
#| fig-height: 5.5

duration_trend <- songs |> 
  group_by(Year) |> 
  summarize(
    Avg_Duration = round(mean(duration_min, na.rm = TRUE), 2),
    Median_Duration = round(median(duration_min, na.rm = TRUE), 2),
    Shortest_Song = Song[which.min(duration_min)],
    Shortest_Artist = Artist[which.min(duration_min)],
    Shortest_Min = round(min(duration_min), 2),
    Longest_Song = Song[which.max(duration_min)],
    Longest_Artist = Artist[which.max(duration_min)],
    Longest_Min = round(max(duration_min), 2),
    Count = n()
  ) |> 
  mutate(
    Tooltip = paste0(
      "<b>Year:</b> ", Year, "<br>",
      "<b>Avg Duration:</b> ", Avg_Duration, " min<br>",
      "<b>Median Duration:</b> ", Median_Duration, " min<br>",
      "<b>Longest Hit:</b> ", Longest_Song, " (", Longest_Min, " min)<br>",
      "<b>Shortest Hit:</b> ", Shortest_Song, " (", Shortest_Min, " min)"
    )
  )

p_duration <- ggplot(duration_trend, aes(x = Year, y = Avg_Duration)) +
  geom_area(fill = "#10b981", alpha = 0.15) +
  geom_line(color = "#059669", linewidth = 1.1) +
  geom_point(aes(text = Tooltip), color = "#047857", size = 2) +
  scale_y_continuous(breaks = seq(2.0, 5.0, by = 0.5), limits = c(2.2, 4.8)) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  labs(
    x = "Year",
    y = "Average Song Duration (Minutes)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_text(color = "#1e293b"),
    panel.grid.minor = element_blank()
  )

ggplotly(p_duration, height = 480, tooltip = "text") |> 
  layout(
    autosize = TRUE,
    dragmode = FALSE,
    xaxis = list(fixedrange = TRUE),
    yaxis = list(fixedrange = TRUE)
  ) |> 
  config(responsive = TRUE, scrollZoom = FALSE, displayModeBar = FALSE)
```

## The Rise of Danceability

Spotify’s **danceability** metric scores how suitable a track is for dancing based on tempo, rhythm stability, beat strength, and overall regularity. As drum machines, club synthesizers, and hip-hop 808s took over top-40 radio, the danceability of Billboard hits increased by over 25%.

::: {.panel-tabset}

### Overall Trend

```{r}
#| label: danceability-overall-chart
#| fig-height: 5.2

dance_overall <- songs |> 
  group_by(Year) |> 
  summarize(
    Avg_Dance = round(mean(danceability, na.rm = TRUE), 3),
    Count = n(),
    Top_Song = Song[which.max(danceability)],
    Top_Artist = Artist[which.max(danceability)],
    Top_Score = round(max(danceability, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) |> 
  mutate(
    Tooltip = paste0(
      "<b>Year:</b> ", Year, "<br>",
      "<b>Avg Danceability:</b> ", round(Avg_Dance * 100, 1), "%<br>",
      "<b>Top Dance Track:</b> ", Top_Song, " (", Top_Artist, ", ", Top_Score, "%)<br>",
      "<b>Hits Analyzed:</b> ", Count
    )
  )

p_dance_overall <- ggplot(dance_overall, aes(x = Year, y = Avg_Dance)) +
  geom_area(fill = "#8b5cf6", alpha = 0.15) +
  geom_line(color = "#7c3aed", linewidth = 1.1) +
  geom_point(aes(text = Tooltip), color = "#6d28d9", size = 2) +
  geom_smooth(method = "loess", color = "#ec4899", linewidth = 0.9, linetype = "dashed", se = FALSE) +
  scale_y_continuous(labels = scales::percent_format(scale = 100), limits = c(0.45, 0.75)) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  labs(
    x = "Year",
    y = "Average Danceability (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_text(color = "#1e293b"),
    panel.grid.minor = element_blank()
  )

ggplotly(p_dance_overall, height = 460, tooltip = "text") |> 
  layout(
    autosize = TRUE,
    dragmode = FALSE,
    xaxis = list(fixedrange = TRUE),
    yaxis = list(fixedrange = TRUE)
  ) |> 
  config(responsive = TRUE, scrollZoom = FALSE, displayModeBar = FALSE)
```

### Genre Comparison

```{r}
#| label: danceability-genre-chart
#| fig-height: 5.2

dance_genre <- songs |> 
  group_by(Year, Genre) |> 
  summarize(
    Avg_Dance = round(mean(danceability, na.rm = TRUE), 3),
    Count = n(),
    .groups = "drop"
  ) |> 
  filter(Count >= 3) |> 
  group_by(Genre) |> 
  arrange(Year) |> 
  mutate(
    Smooth_Dance = stats::filter(Avg_Dance, rep(1/3, 3), sides = 2),
    Smooth_Dance = ifelse(is.na(Smooth_Dance), Avg_Dance, Smooth_Dance),
    Tooltip = paste0(
      "<b>Genre:</b> ", Genre, "<br>",
      "<b>Year:</b> ", Year, "<br>",
      "<b>Avg Danceability:</b> ", round(Smooth_Dance * 100, 1), "%<br>",
      "<b>Hits in Year:</b> ", Count
    )
  ) |> 
  ungroup()

p_dance_genre <- ggplot(dance_genre, aes(x = Year, y = Smooth_Dance, color = Genre, group = Genre, text = Tooltip)) +
  geom_line(linewidth = 1.2, alpha = 0.85) +
  scale_color_manual(values = genre_colors) +
  scale_y_continuous(labels = scales::percent_format(scale = 100), limits = c(0.45, 0.82)) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  labs(
    x = "Year",
    y = "Danceability (%)",
    color = "Genre"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_text(color = "#1e293b"),
    panel.grid.minor = element_blank()
  )

ggplotly(p_dance_genre, height = 460, tooltip = "text") |> 
  layout(
    autosize = TRUE,
    dragmode = FALSE,
    xaxis = list(fixedrange = TRUE),
    yaxis = list(fixedrange = TRUE)
  ) |> 
  config(responsive = TRUE, scrollZoom = FALSE, displayModeBar = FALSE)
```

:::

## Song Audio Profile Explorer

Explore, compare, and listen to hits across 67 years of chart history. Filter by genre, search your favorite artist, and inspect visual audio meters for danceability, energy, and mood:

::: {.panel-tabset}

### Interactive Audio Matrix

Hover over any hit to inspect its sonic fingerprint (tempo, danceability, energy, and mood). Use the genre legend to isolate musical styles or zoom into specific tempo ranges:

```{r}
#| label: audio-scatter-matrix
#| fig-height: 5.5

p_scatter <- ggplot(songs, aes(
  x = tempo,
  y = danceability,
  color = Genre,
  size = energy,
  text = paste0(
    "<b>", Song, "</b> (", Year, ")<br>",
    "<b>Artist:</b> ", Artist, "<br>",
    "<b>Genre:</b> ", Genre, "<br>",
    "<b>Tempo:</b> ", round(tempo), " BPM<br>",
    "<b>Danceability:</b> ", round(danceability * 100, 1), "%<br>",
    "<b>Energy:</b> ", round(energy * 100, 1), "%<br>",
    "<b>Mood (Valence):</b> ", round(valence * 100, 1), "%"
  )
)) +
  geom_point(alpha = 0.55) +
  scale_color_manual(values = genre_colors) +
  scale_y_continuous(labels = scales::percent_format(scale = 100), limits = c(0.2, 1.0)) +
  scale_x_continuous(limits = c(50, 220), breaks = seq(60, 200, by = 20)) +
  scale_size_continuous(range = c(1.5, 4.5), guide = "none") +
  labs(
    x = "Tempo / BPM (Beats Per Minute)",
    y = "Danceability (%)",
    color = "Genre"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_text(color = "#1e293b"),
    panel.grid.minor = element_blank()
  )

ggplotly(p_scatter, height = 480, tooltip = "text") |> 
  layout(
    autosize = TRUE,
    legend = list(orientation = "h", x = 0.5, xanchor = "center", y = 1.1)
  ) |> 
  config(responsive = TRUE, scrollZoom = FALSE, displayModeBar = FALSE)
```

### Searchable Song Vault & Audio Meters

```{r}
#| label: audio-table

songs_table <- songs |> 
  select(
    Year,
    Song,
    Artist,
    Genre,
    Duration = duration_min,
    BPM = tempo,
    Dance = danceability,
    Energy = energy,
    Mood = valence
  ) |> 
  mutate(
    BPM = round(BPM),
    Duration = paste0(round(Duration, 1), "m")
  )

reactable(
  songs_table,
  filterable = TRUE,
  searchable = TRUE,
  defaultPageSize = 10,
  pageSizeOptions = c(10, 25, 50),
  showPageSizeOptions = TRUE,
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  columns = list(
    Year = colDef(width = 70, align = "center"),
    Song = colDef(
      minWidth = 170,
      cell = JS("function(cellInfo) {
        var song = cellInfo.value;
        var artist = cellInfo.row['Artist'];
        var q = encodeURIComponent(song + ' ' + artist);
        var url = 'https://www.youtube.com/results?search_query=' + q;
        return '<div style=\"display:flex;align-items:center;justify-content:space-between;\">' +
               '<span style=\"font-weight:600;\">' + song + '</span>' +
               '<a href=\"' + url + '\" target=\"_blank\" rel=\"noopener noreferrer\" title=\"Listen on YouTube\" ' +
               'style=\"padding:2px 7px;background:#fee2e2;color:#dc2626;border-radius:4px;font-weight:600;font-size:0.75rem;text-decoration:none;margin-left:6px;\">▶</a>' +
               '</div>';
      }"),
      html = TRUE
    ),
    Artist = colDef(minWidth = 140),
    Genre = colDef(
      width = 135,
      cell = JS("function(cellInfo) {
        var colors = {
          'Rock': '#ef4444',
          'Pop': '#ec4899',
          'Hip-Hop/Rap': '#2563eb',
          'R&B/Soul': '#8b5cf6',
          'Country': '#f59e0b',
          'Electronic/Dance': '#06b6d4'
        };
        var bg = colors[cellInfo.value] || '#64748b';
        return '<span style=\"background-color:' + bg + ';color:#ffffff;padding:2px 8px;border-radius:12px;font-size:0.75rem;font-weight:600;\">' + cellInfo.value + '</span>';
      }"),
      html = TRUE
    ),
    Duration = colDef(width = 75, align = "right"),
    BPM = colDef(width = 75, align = "right"),
    Dance = colDef(
      width = 120,
      cell = JS("function(cellInfo) {
        var pct = Math.round(cellInfo.value * 100);
        return '<div style=\"display:flex;align-items:center;font-family:monospace;font-size:0.82rem;\">' +
               '<div style=\"flex-grow:1;background:#f1f5f9;height:12px;border-radius:2px;overflow:hidden;margin-right:8px;\">' +
               '<div style=\"width:' + pct + '%;height:100%;background:#10b981;border-radius:2px;\"></div></div>' +
               '<span>' + pct + '%</span></div>';
      }"),
      html = TRUE
    ),
    Energy = colDef(
      width = 120,
      cell = JS("function(cellInfo) {
        var pct = Math.round(cellInfo.value * 100);
        return '<div style=\"display:flex;align-items:center;font-family:monospace;font-size:0.82rem;\">' +
               '<div style=\"flex-grow:1;background:#f1f5f9;height:12px;border-radius:2px;overflow:hidden;margin-right:8px;\">' +
               '<div style=\"width:' + pct + '%;height:100%;background:#f59e0b;border-radius:2px;\"></div></div>' +
               '<span>' + pct + '%</span></div>';
      }"),
      html = TRUE
    ),
    Mood = colDef(
      width = 120,
      cell = JS("function(cellInfo) {
        var pct = Math.round(cellInfo.value * 100);
        return '<div style=\"display:flex;align-items:center;font-family:monospace;font-size:0.82rem;\">' +
               '<div style=\"flex-grow:1;background:#f1f5f9;height:12px;border-radius:2px;overflow:hidden;margin-right:8px;\">' +
               '<div style=\"width:' + pct + '%;height:100%;background:#3b82f6;border-radius:2px;\"></div></div>' +
               '<span>' + pct + '%</span></div>';
      }"),
      html = TRUE
    )
  ),
  theme = reactableTheme(
    borderColor = "#e2e8f0",
    stripedColor = "#f8fafc",
    highlightColor = "#eff6ff",
    cellPadding = "8px 12px",
    style = list(fontFamily = "-apple-system, BlinkMacSystemFont, Segoe UI, Roboto, sans-serif", fontSize = "0.88rem")
  )
)
```

:::
