---
title: "Anatomy of a Hit"
subtitle: "Interactive exploration of song length, danceability, tempo, and acousticness across 5,800+ Billboard hits."
format:
  html:
    toc: true
    page-layout: article
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(plotly)
library(reactable)

songs <- read_csv("data/billboard_audio_features.csv", show_col_types = FALSE)

genre_colors <- c(
  "Rock" = "#ef4444",
  "Pop" = "#ec4899",
  "Hip-Hop/Rap" = "#2563eb",
  "R&B/Soul" = "#8b5cf6",
  "Country" = "#f59e0b",
  "Electronic/Dance" = "#06b6d4"
)
```

## The Shrinking Song: Duration Across Decades

Why are modern songs getting shorter? In the 1960s, 7-inch vinyl records limited songs to under 3 minutes. The CD revolution in the 1980s and 1990s unlocked epic 5-minute anthems. But since 2015, streaming royalty models (which pay per play once a listener passes 30 seconds) and short-form video on TikTok have pushed hit songs back down to historic lows.

```{r}
#| label: song-duration-chart
#| fig-height: 5.5

duration_trend <- songs |> 
  group_by(Year) |> 
  summarize(
    Avg_Duration = round(mean(duration_min, na.rm = TRUE), 2),
    Median_Duration = round(median(duration_min, na.rm = TRUE), 2),
    Shortest_Song = Song[which.min(duration_min)],
    Shortest_Artist = Artist[which.min(duration_min)],
    Shortest_Min = round(min(duration_min), 2),
    Longest_Song = Song[which.max(duration_min)],
    Longest_Artist = Artist[which.max(duration_min)],
    Longest_Min = round(max(duration_min), 2),
    Count = n()
  ) |> 
  mutate(
    Tooltip = paste0(
      "<b>Year:</b> ", Year, "<br>",
      "<b>Avg Duration:</b> ", Avg_Duration, " min<br>",
      "<b>Median Duration:</b> ", Median_Duration, " min<br>",
      "<b>Longest Hit:</b> ", Longest_Song, " (", Longest_Min, " min)<br>",
      "<b>Shortest Hit:</b> ", Shortest_Song, " (", Shortest_Min, " min)"
    )
  )

p_duration <- ggplot(duration_trend, aes(x = Year, y = Avg_Duration)) +
  geom_area(fill = "#10b981", alpha = 0.15) +
  geom_line(color = "#059669", linewidth = 1.1) +
  geom_point(aes(text = Tooltip), color = "#047857", size = 2) +
  scale_y_continuous(breaks = seq(2.0, 5.0, by = 0.5), limits = c(2.2, 4.8)) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  labs(
    x = "Year",
    y = "Average Song Duration (Minutes)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_text(color = "#1e293b"),
    panel.grid.minor = element_blank()
  )

ggplotly(p_duration, height = 480, tooltip = "text") |> 
  layout(
    autosize = TRUE,
    dragmode = FALSE,
    xaxis = list(fixedrange = TRUE),
    yaxis = list(fixedrange = TRUE)
  ) |> 
  config(responsive = TRUE, scrollZoom = FALSE, displayModeBar = FALSE)
```

## The Rise of Danceability

Spotify’s **danceability** metric scores how suitable a track is for dancing based on tempo, rhythm stability, beat strength, and overall regularity. As drum machines, club synthesizers, and hip-hop 808s took over top-40 radio, the danceability of Billboard hits increased by over 25%.

```{r}
#| label: danceability-chart
#| fig-height: 5.5

dance_trend <- songs |> 
  group_by(Year, Genre) |> 
  summarize(
    Avg_Dance = round(mean(danceability, na.rm = TRUE), 3),
    Count = n(),
    .groups = "drop"
  ) |> 
  filter(Count >= 3) |> 
  mutate(
    Tooltip = paste0(
      "<b>Year:</b> ", Year, "<br>",
      "<b>Genre:</b> ", Genre, "<br>",
      "<b>Avg Danceability:</b> ", round(Avg_Dance * 100, 1), "%<br>",
      "<b>Hits in Year:</b> ", Count
    )
  )

p_dance <- ggplot(dance_trend, aes(x = Year, y = Avg_Dance, color = Genre)) +
  geom_line(linewidth = 0.9, alpha = 0.85) +
  geom_point(aes(text = Tooltip), size = 1.8) +
  scale_color_manual(values = genre_colors) +
  scale_y_continuous(labels = scales::percent_format(scale = 100), limits = c(0.4, 0.85)) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  labs(
    x = "Year",
    y = "Danceability (%)",
    color = "Genre"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_text(color = "#1e293b"),
    panel.grid.minor = element_blank()
  )

ggplotly(p_dance, height = 480, tooltip = "text") |> 
  layout(
    autosize = TRUE,
    dragmode = FALSE,
    xaxis = list(fixedrange = TRUE),
    yaxis = list(fixedrange = TRUE)
  ) |> 
  config(responsive = TRUE, scrollZoom = FALSE, displayModeBar = FALSE)
```

## Song Audio Profile Explorer

Search, sort, and inspect audio attributes for any hit song in the catalog. Filter by genre, BPM, or danceability score:

```{r}
#| label: audio-table

songs_table <- songs |> 
  select(
    Year,
    Song,
    Artist,
    Genre,
    Duration = duration_min,
    BPM = tempo,
    Dance = danceability,
    Energy = energy,
    Mood = valence
  ) |> 
  mutate(
    Dance = paste0(round(Dance * 100, 0), "%"),
    Energy = paste0(round(Energy * 100, 0), "%"),
    Mood = paste0(round(Mood * 100, 0), "%"),
    Duration = paste0(Duration, " min")
  )

reactable(
  songs_table,
  filterable = TRUE,
  searchable = TRUE,
  defaultPageSize = 10,
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  columns = list(
    Year = colDef(width = 75, align = "center"),
    Song = colDef(minWidth = 160, style = list(fontWeight = "bold")),
    Artist = colDef(minWidth = 150),
    Genre = colDef(width = 130),
    Duration = colDef(width = 100, align = "right"),
    BPM = colDef(width = 80, align = "right"),
    Dance = colDef(width = 85, align = "right"),
    Energy = colDef(width = 85, align = "right"),
    Mood = colDef(width = 85, align = "right")
  ),
  theme = reactableTheme(
    borderColor = "#e2e8f0",
    stripedColor = "#f8fafc",
    highlightColor = "#eff6ff",
    cellPadding = "8px 12px",
    style = list(fontFamily = "-apple-system, BlinkMacSystemFont, Segoe UI, Roboto, sans-serif", fontSize = "0.88rem")
  )
)
```
