---
title: "What does music say about the world?"
subtitle: "A visual history of Billboard Hot 100 genres and hits (1958–2024)"
page-layout: full
toc: false
execute:
  echo: false
---

Popular music offers a lens into cultural moments, social shifts, and technological revolutions. By tracking which genres dominated the Billboard charts over seven decades, we can see how American musical tastes have transformed over time.

```{r setup}
#| message: false
library(tidyverse)
library(plotly)

billboard_all_time <- read_csv("data/billboard_all_time_top_100.csv")
billboard_history <- read_csv("data/billboard_historical_genres.csv")

genre_colors <- c(
  "Rock" = "#ef4444",
  "Pop" = "#ec4899",
  "R&B/Soul" = "#8b5cf6",
  "Hip-Hop/Rap" = "#2563eb",
  "Country" = "#f59e0b",
  "Electronic/Dance" = "#06b6d4"
)
```

## Every Single Hit Song (1958 – 2024)

Explore all 6,700 year-end hits across chart history. Hover over any point to see the song title, artist, year-end rank, and peak chart position.

```{r}
#| message: false
#| out-width: "100%"
#| fig-height: 7

p_all_songs <- ggplot(billboard_history, aes(
  x = Year,
  y = Year_Rank,
  color = Genre,
  text = paste0(
    "<b>", Song, "</b><br>",
    "Artist: ", Artist, "<br>",
    "Year: ", Year, " (Rank #", Year_Rank, ")<br>",
    "Peak: #", Peak, "<br>",
    "Weeks Charted: ", Weeks, "<br>",
    "Genre: ", Genre
  )
)) +
  geom_point(alpha = 0.65, size = 1.2) +
  # Reverse y-axis so rank #1 is at the top
  scale_y_reverse(breaks = c(1, 25, 50, 75, 100), limits = c(101, 0)) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 5)) +
  scale_color_manual(values = genre_colors) +
  labs(
    title = NULL,
    x = "Year",
    y = "Year-End Rank (1 = Top Hit)",
    color = "Genre"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_text(color = "#1e293b"),
    panel.grid.minor = element_blank()
  )

ggplotly(p_all_songs, height = 550, tooltip = "text") |> 
  layout(autosize = TRUE) |> 
  config(responsive = TRUE)
```

## Billboard Greatest of All Time: Genre Breakdown

```{r}
#| message: false
#| out-width: "100%"
#| fig-height: 7

genre_stats <- billboard_all_time |> 
  group_by(Genre) |> 
  summarize(
    Count = n(),
    Avg_Rank = round(mean(All_Time_Rank), 1),
    Best_Rank = min(All_Time_Rank),
    Top_Song = Song[which.min(All_Time_Rank)],
    Top_Cover = Cover[which.min(All_Time_Rank)],
    Top_Artist = Artist[which.min(All_Time_Rank)],
    Worst_Rank = max(All_Time_Rank),
    Lowest_Song = Song[which.max(All_Time_Rank)],
    Lowest_Artist = Artist[which.max(All_Time_Rank)],
    .groups = "drop"
  ) |> 
  mutate(Genre_Fct = fct_reorder(Genre, Count)) |> 
  arrange(Genre_Fct)

p_all_time <- ggplot(genre_stats, aes(
  x = Count, 
  y = Genre_Fct,
  fill = Avg_Rank,
  text = paste0(
    "<b>", Genre, "</b><br>",
    "Songs in All-Time Top 100: ", Count, "<br>",
    "Average All-Time Rank: #", Avg_Rank, "<br>",
    "Highest Hit: #", Best_Rank, " ", Top_Song, " (", Top_Artist, ")"
  )
)) +
  geom_col(width = 0.72, color = "white", linewidth = 0.6, alpha = 0.92) +
  geom_text(
    aes(x = Count / 2, label = ifelse(Count >= 4, paste0(Count, " songs"), as.character(Count))), 
    color = "white", 
    fontface = "bold", 
    size = 3.6
  ) +
  scale_fill_gradientn(
    colors = c("#06b6d4", "#3b82f6", "#6366f1", "#a855f7", "#ec4899", "#f43f5e"),
    name = "Avg Rank\n(#1 is top)"
  ) +
  scale_x_continuous(limits = c(0, 32), expand = expansion(mult = c(0, 0.03))) +
  labs(
    title = NULL,
    x = "Number of Songs in All-Time Top 100",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", size = 11, color = "#1e293b"),
    axis.text.x = element_text(color = "#64748b")
  )

img_list <- lapply(seq_len(nrow(genre_stats)), \(i) {
  list(
    source = genre_stats$Top_Cover[i],
    xref = "x",
    yref = "y",
    x = genre_stats$Count[i] + 0.6,
    y = i,
    sizex = 1.3,
    sizey = 0.72,
    xanchor = "left",
    yanchor = "middle",
    sizing = "contain"
  )
})

annot_list <- lapply(seq_len(nrow(genre_stats)), \(i) {
  list(
    xref = "x",
    yref = "y",
    x = genre_stats$Count[i] + 2.2,
    y = i,
    text = paste0("#", genre_stats$Best_Rank[i], " ", genre_stats$Top_Song[i]),
    showarrow = FALSE,
    xanchor = "left",
    yanchor = "middle",
    font = list(size = 11, color = "#1e293b")
  )
})

ggplotly(p_all_time, height = 540, tooltip = "text") |> 
  layout(
    autosize = TRUE,
    images = img_list, 
    annotations = annot_list
  ) |> 
  config(responsive = TRUE)
```

## Genre Evolution: Annual Billboard Top 100 (1958 – 2024)

```{r}
#| message: false
#| out-width: "100%"
#| fig-height: 7

genre_summary <- billboard_history |> 
  group_by(Year, Genre) |> 
  summarize(
    Count = n(),
    Top_Song = Song[which.min(Peak)],
    Top_Artist = Artist[which.min(Peak)],
    Best_Peak = min(Peak),
    Lowest_Song = Song[which.max(Peak)],
    Lowest_Artist = Artist[which.max(Peak)],
    Worst_Peak = max(Peak),
    .groups = "drop_last"
  ) |> 
  mutate(
    Percent = round((Count / sum(Count)) * 100, 1),
    Tooltip = ifelse(
      Count == 1,
      paste0(
        "Year: ", Year, "<br>",
        "Genre: ", Genre, "<br>",
        "Share: ", Percent, "% (1 song)<br>",
        "Top Hit: #", Best_Peak, " ", Top_Song, " (", Top_Artist, ")"
      ),
      paste0(
        "Year: ", Year, "<br>",
        "Genre: ", Genre, "<br>",
        "Share: ", Percent, "% (", Count, " songs)<br>",
        "Top Hit: #", Best_Peak, " ", Top_Song, " (", Top_Artist, ")"
      )
    )
  ) |> 
  ungroup()

p <- ggplot(genre_summary, aes(
  x = Year,
  y = Percent,
  fill = Genre,
  text = Tooltip
)) +
  geom_col(position = "stack", width = 0.85, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = genre_colors) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 5), expand = expansion(mult = c(0.01, 0.01))) +
  labs(
    title = NULL,
    x = "Year",
    y = "Share of Top Hits",
    fill = "Genre"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(face = "bold", size = 10, color = "#1e293b"),
    axis.text.y = element_text(color = "#64748b"),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

ggplotly(p, height = 550, tooltip = "text") |> 
  layout(autosize = TRUE) |> 
  config(responsive = TRUE)
```
