---
title: "Explore the Data"
subtitle: "Billboard Hot 100 historical records and project overview"
format:
  html:
    theme: cosmo
    page-layout: full
    toc: false
execute:
  echo: false
---


::: {.callout-note icon=false}
### Project Plan

I am planning to compare the top genres in songs from the Billboard Top 100 from the beginning and median years of each decade since the Billboard Top 100 has started. I am also going to find different impactful events from these years or surrounding years and see the impact on the popular genres during that time.
:::

Below is the complete song-level dataset used in this project. You can sort, filter, or search for any artist, song, or genre across the tables below.


```{r}
#| message: false
library(tidyverse)
library(gt)

billboard_all_time <- read_csv("data/billboard_all_time_top_100.csv")
billboard_history <- read_csv("data/billboard_historical_genres.csv")
```

## Billboard Greatest of All Time Hot 100 Songs

```{r}
billboard_all_time |> 
  select(All_Time_Rank, Cover, Song, Artist, Genre, Year, Peak, Weeks, Weeks_at_1) |> 
  gt() |> 
  tab_header(
    title = "Billboard Greatest of All Time Hot 100",
    subtitle = "Official ranking and chart statistics of the top songs in Billboard history"
  ) |> 
  cols_label(
    All_Time_Rank = "#",
    Cover = "",
    Song = "Song Title",
    Artist = "Artist",
    Genre = "Genre",
    Year = "Year",
    Peak = "Peak Rank",
    Weeks = "Weeks Charted",
    Weeks_at_1 = "Weeks at #1"
  ) |> 
  fmt_image(columns = Cover, height = 45) |> 
  tab_options(table.width = pct(100)) |> 
  opt_interactive(
    use_pagination = TRUE, 
    page_size_default = 25, 
    use_page_size_select = TRUE,
    page_size_values = c(10, 25, 50, 100),
    use_search = TRUE
  )
```

## Billboard Hot 100: Every Year (1958 – 2024)

```{r}
billboard_history |> 
  select(Year, Year_Rank, Cover, Song, Artist, Genre, Peak, Weeks) |> 
  gt() |> 
  tab_header(
    title = "Billboard Top 100 Annual Hits (1958 – 2024)",
    subtitle = "Top 100 songs from every year of the Billboard Hot 100 chart history"
  ) |> 
  cols_label(
    Year = "Year",
    Year_Rank = "Rank",
    Cover = "",
    Song = "Song",
    Artist = "Artist",
    Genre = "Genre",
    Peak = "Peak Rank",
    Weeks = "Weeks Charted"
  ) |> 
  fmt_image(columns = Cover, height = 45) |> 
  tab_options(table.width = pct(100)) |> 
  opt_interactive(
    use_pagination = TRUE, 
    page_size_default = 25, 
    use_page_size_select = TRUE,
    page_size_values = c(10, 25, 50, 100),
    use_search = TRUE
  )
```

