---
title: "Project Overview & Methodology"
subtitle: "Data provenance, analytical methodology, and complete Billboard Hot 100 historical records"
format:
  html:
    page-layout: full
    toc: false
execute:
  echo: false
---

::: {.profile-card}
### Research Lead & Affiliation

::: {.profile-grid}
::: {.profile-item}
::: {.profile-item-label}
Principal Investigator
:::
::: {.profile-item-value}
Celise Cupit
:::
:::

::: {.profile-item}
::: {.profile-item-label}
Academic Institution
:::
::: {.profile-item-value}
[New College of Florida](https://www.ncf.edu/)
:::
:::

::: {.profile-item}
::: {.profile-item-label}
Contact Email
:::
::: {.profile-item-value}
[c.cupit30@ncf.edu](mailto:c.cupit30@ncf.edu)
:::
:::

::: {.profile-item}
::: {.profile-item-label}
Project Repository
:::
::: {.profile-item-value}
[github.com/cecupit4/mind-over-music](https://github.com/ccupit4/mind-over-music)
:::
:::
:::
:::

::: {.callout-note icon=false}
### Analytical Scope & Objectives
This research project analyzes longitudinal shifts in musical consumption by evaluating genre share across the Billboard Hot 100 from 1958 to 2024. The study examines benchmark years and pivotal transition eras to quantify how macro-cultural events, recording technology disruptions (vinyl to cassette, CD, digital download, and on-demand streaming), and demographic changes influenced the acoustic and emotional profile of mainstream hits.
:::

Below is the verified song-level database supporting this analysis. Use the search and pagination controls to filter by artist, genre, release year, or peak chart performance.


```{r}
#| message: false
library(tidyverse)
library(gt)

billboard_all_time <- read_csv("data/billboard_all_time_top_100.csv")
billboard_history <- read_csv("data/billboard_historical_genres.csv")
```

## Billboard Greatest of All Time Hot 100 Songs

```{r}
billboard_all_time |> 
  select(All_Time_Rank, Cover, Song, Artist, Genre, Year, Peak, Weeks, Weeks_at_1) |> 
  gt() |> 
  tab_header(
    title = "Billboard Greatest of All Time Hot 100",
    subtitle = "Official ranking and chart statistics of the top songs in Billboard history"
  ) |> 
  cols_label(
    All_Time_Rank = "#",
    Cover = "",
    Song = "Song Title",
    Artist = "Artist",
    Genre = "Genre",
    Year = "Year",
    Peak = "Peak Rank",
    Weeks = "Weeks Charted",
    Weeks_at_1 = "Weeks at #1"
  ) |> 
  fmt_image(columns = Cover, height = 45) |> 
  tab_options(table.width = pct(100)) |> 
  opt_interactive(
    use_pagination = TRUE, 
    page_size_default = 25, 
    use_page_size_select = TRUE,
    page_size_values = c(10, 25, 50, 100),
    use_search = TRUE
  )
```

## Billboard Hot 100: Every Year (1958 – 2024)

```{r}
billboard_history |> 
  select(Year, Year_Rank, Cover, Song, Artist, Genre, Peak, Weeks) |> 
  gt() |> 
  tab_header(
    title = "Billboard Top 100 Annual Hits (1958 – 2024)",
    subtitle = "Top 100 songs from every year of the Billboard Hot 100 chart history"
  ) |> 
  cols_label(
    Year = "Year",
    Year_Rank = "Rank",
    Cover = "",
    Song = "Song",
    Artist = "Artist",
    Genre = "Genre",
    Peak = "Peak Rank",
    Weeks = "Weeks Charted"
  ) |> 
  fmt_image(columns = Cover, height = 45) |> 
  tab_options(table.width = pct(100)) |> 
  opt_interactive(
    use_pagination = TRUE, 
    page_size_default = 25, 
    use_page_size_select = TRUE,
    page_size_values = c(10, 25, 50, 100),
    use_search = TRUE
  )
```

