---
title: "The Emotional Arc of Pop"
subtitle: "Are hit songs actually getting sadder? Six decades of musical mood and sentiment on the Billboard Hot 100."
format:
  html:
    toc: true
    page-layout: article
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(plotly)

songs <- read_csv("data/billboard_audio_features.csv", show_col_types = FALSE)

genre_colors <- c(
  "Rock" = "#ef4444",
  "Pop" = "#ec4899",
  "Hip-Hop/Rap" = "#2563eb",
  "R&B/Soul" = "#8b5cf6",
  "Country" = "#f59e0b",
  "Electronic/Dance" = "#06b6d4"
)
```

## The Great Sadness Shift: 1958 to 2024

In music theory and audio analytics, **valence** measures musical positivity. Tracks with high valence sound cheerful, euphoric, and upbeat (like 1960s Motown and 1970s disco), while tracks with low valence sound melancholic, moody, or angry (like 1990s grunge and modern lo-fi pop).

Over the last 60 years, the average emotional positivity of hit music has declined steadily. In the early 1960s, the average hit had a valence score around **0.65**. By the 2020s, that average plummeted to **0.49**.

```{r}
#| label: annual-valence-trend
#| fig-height: 5.5

annual_mood <- songs |> 
  group_by(Year) |> 
  summarize(
    Avg_Valence = round(mean(valence, na.rm = TRUE), 3),
    Avg_Energy = round(mean(energy, na.rm = TRUE), 3),
    Count = n(),
    Happiest_Song = Song[which.max(valence)],
    Happiest_Artist = Artist[which.max(valence)],
    Saddest_Song = Song[which.min(valence)],
    Saddest_Artist = Artist[which.min(valence)]
  ) |> 
  mutate(
    Tooltip = paste0(
      "<b>Year:</b> ", Year, "<br>",
      "<b>Avg Positivity (Valence):</b> ", round(Avg_Valence * 100, 1), "%<br>",
      "<b>Avg Energy:</b> ", round(Avg_Energy * 100, 1), "%<br>",
      "<b>Happiest Hit:</b> ", Happiest_Song, " (", Happiest_Artist, ")<br>",
      "<b>Moodiest Hit:</b> ", Saddest_Song, " (", Saddest_Artist, ")"
    )
  )

p_valence <- ggplot(annual_mood, aes(x = Year, y = Avg_Valence)) +
  geom_area(fill = "#3b82f6", alpha = 0.15) +
  geom_line(color = "#1d4ed8", linewidth = 1.1) +
  geom_point(aes(text = Tooltip), color = "#2563eb", size = 2) +
  geom_smooth(method = "loess", color = "#dc2626", linewidth = 0.8, linetype = "dashed", se = FALSE) +
  scale_y_continuous(labels = scales::percent_format(scale = 100), limits = c(0.4, 0.75)) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  labs(
    x = "Year",
    y = "Average Positivity / Valence (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_text(color = "#1e293b"),
    panel.grid.minor = element_blank()
  )

ggplotly(p_valence, height = 480, tooltip = "text") |> 
  layout(
    autosize = TRUE,
    dragmode = FALSE,
    xaxis = list(fixedrange = TRUE),
    yaxis = list(fixedrange = TRUE)
  ) |> 
  config(responsive = TRUE, scrollZoom = FALSE, displayModeBar = FALSE)
```

::: {.callout-note}
### Why Did Hits Get Sadder?
Musicologists point to three driving factors:
1. **The Intimacy of Headphones:** The Walkman in the 1980s, the iPod in the 2000s, and Spotify smartphones turned music from a communal party experience into a private, introspective soundtrack for loneliness and emotion.
2. **Minor Key Dominance:** In the 1960s, over 85% of Billboard hits were written in major keys. Today, more than 40% are composed in minor scales, signaling tension and vulnerability.
3. **The Rise of Trap & Emo-Rap:** Modern chart-toppers from artists like Juice WRLD, Billie Eilish, and Olivia Rodrigo embrace vulnerability, mental health, and heartbreak.
:::

## The Mood Matrix: Energy vs. Positivity by Genre

How do different genres compare in emotional space? When we plot **Energy** against **Positivity (Valence)**, four distinct musical quadrants emerge:

- **Top-Right (High Energy, High Positivity):** Joyful party anthems and dancefloor euphoria.
- **Top-Left (High Energy, Low Positivity):** Aggressive rock, metal, and intense club beats.
- **Bottom-Right (Low Energy, High Positivity):** Warm, peaceful acoustic ballads and nostalgic love songs.
- **Bottom-Left (Low Energy, Low Positivity):** Melancholic, chill, and sorrowful tracks.

```{r}
#| label: mood-quadrant
#| fig-height: 6

genre_mood <- songs |> 
  group_by(Genre) |> 
  summarize(
    Avg_Valence = round(mean(valence, na.rm = TRUE), 3),
    Avg_Energy = round(mean(energy, na.rm = TRUE), 3),
    Count = n(),
    Top_Example = Song[1],
    Top_Artist = Artist[1]
  ) |> 
  mutate(
    Tooltip = paste0(
      "<b>Genre:</b> ", Genre, "<br>",
      "<b>Positivity (Valence):</b> ", round(Avg_Valence * 100, 1), "%<br>",
      "<b>Energy:</b> ", round(Avg_Energy * 100, 1), "%<br>",
      "<b>Catalog Songs:</b> ", Count, "<br>",
      "<b>Example Hit:</b> ", Top_Example, " (", Top_Artist, ")"
    )
  )

p_quad <- ggplot(genre_mood, aes(x = Avg_Valence, y = Avg_Energy, color = Genre)) +
  geom_hline(yintercept = 0.62, linetype = "dashed", color = "#cbd5e1", linewidth = 0.6) +
  geom_vline(xintercept = 0.58, linetype = "dashed", color = "#cbd5e1", linewidth = 0.6) +
  geom_point(aes(text = Tooltip), size = 6, alpha = 0.85) +
  scale_color_manual(values = genre_colors) +
  scale_x_continuous(labels = scales::percent_format(scale = 100), limits = c(0.45, 0.7)) +
  scale_y_continuous(labels = scales::percent_format(scale = 100), limits = c(0.5, 0.75)) +
  labs(
    x = "Positivity / Valence (%)",
    y = "Energy Level (%)",
    color = "Genre"
  ) +
  annotate("text", x = 0.68, y = 0.74, label = "🎉 Joyful & Upbeat", color = "#64748b", size = 3.5, fontface = "bold", hjust = 1) +
  annotate("text", x = 0.46, y = 0.74, label = "⚡ Intense & Aggressive", color = "#64748b", size = 3.5, fontface = "bold", hjust = 0) +
  annotate("text", x = 0.68, y = 0.51, label = "☀️ Warm & Calm", color = "#64748b", size = 3.5, fontface = "bold", hjust = 1) +
  annotate("text", x = 0.46, y = 0.51, label = "🌧️ Moody & Melancholic", color = "#64748b", size = 3.5, fontface = "bold", hjust = 0) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_text(color = "#1e293b"),
    panel.grid.minor = element_blank()
  )

ggplotly(p_quad, height = 500, tooltip = "text") |> 
  layout(
    autosize = TRUE,
    dragmode = FALSE,
    xaxis = list(fixedrange = TRUE),
    yaxis = list(fixedrange = TRUE)
  ) |> 
  config(responsive = TRUE, scrollZoom = FALSE, displayModeBar = FALSE)
```
